% !TeX root = ../main.tex
% Add the above to each chapter to make compiling the PDF easier in some editors.

\chapter{Introduction}\label{chapter:introduction}
Deep learning has become an important tool across science and engineering, with one of its strongest impacts in computer vision. Today, models perform very well on tasks such as image classification, object detection, and segmentation~\cite{liu2021swin}. These results, however, usually depend on large labeled datasets. In medical imaging, building comprehensive labeled datasets remains difficult, especially in 3D settings, and this has limited progress in several downstream tasks ~\cite{hamamci2024foundation}.

Within medical imaging, radiology report generation is a high-impact task because reports are the main communication channel between image analysis and clinical decision-making~\cite{wang2025rrgsurvey}. A useful model must do more than produce fluent text. It must describe findings accurately, preserve anatomical context, and avoid clinically misleading statements~\cite{ostmeier2024green}. This makes the problem closer to structured clinical reasoning than to generic image captioning.

\section{Problem Statement}
Traditional encoder-decoder approaches for radiology report generation were limited by imperfect image-text alignment and by relatively generic language-generation components~\cite{chen2020generating}. While early 2D Vision-Language Models (VLMs) leveraging attention mechanisms improved the alignment between image regions and text tokens~\cite{chen2022crossmodal}, they were limited in capturing full volumetric context. In clinical practice, critical findings such as lesion shape, spatial extent, and inter-slice continuity often depend on volumetric context~\cite{hatamizadeh2022unetr}. 

Applying modern VLM architectures directly to 3D medical volumes, like CT scans, introduces substantial computational bottlenecks~\cite{hamamci2024ct2rep,hamamci2025bettertokensbetter3d}. Volumetric inputs drastically increase visual token counts, which makes global attention expensive and motivates bridging architectures such as perceptual resamplers or Q-Formers to compress dense visual tokens before decoding~\cite{alayrac2022flamingo,li2023blip2}. Furthermore, while the field of generative \ac{AI} has seen an expansion of capable multi-billion parameter foundation models, recent 3D report-generation studies suggest that both vision-side representation quality and decoder design matter, but their relative contribution remains unclear~\cite{argus2024}.

\section{Thesis Formulation}
This thesis builds on fVLM~\cite{shui2025fvlm} and focuses on radiology report generation from 3D chest CT. Rather than introducing a new vision backbone, we reuse the pretrained fVLM vision encoder and study how to turn its visual features into clinically reasonable reports.

Our main objective is to improve report quality through design choices around this encoder. Since fVLM does not release report-generation weights and provides only limited detail about the decoder-side setup, we construct a reproducible internal reference branch guided by the published description~\cite{shui2025fvlm}. Concretely, this branch uses a \ac{BERTbase} decoder~\cite{devlin2019bert}. We train this branch with language modeling loss on \ac{CTRATE} under our internal protocol (Chapter~\ref{chapter:methodology}). We then evaluate additional generation strategies, including organ-conditioned token interfaces and decoder adaptation. Experiments are conducted on the \ac{CTRATE} dataset \cite{hamamci2024foundation}, a large-scale public dataset pairing 3D chest CT volumes with free-text radiology reports.

Instead of relying only on text-overlap metrics such as BLEU \cite{papineni2002bleu} and ROUGE \cite{lin2004rouge}, we also evaluate clinical quality using \ac{RadGraph} \cite{jain2021radgraph} and \ac{GREEN} \cite{ostmeier2024green}, the Generative Radiology Report Evaluation and Error Notation metric. This helps us judge whether the generated reports are clinically useful, not only lexically similar.

We analyze which decoder choices and interface designs best convert pretrained 3D visual representations into structured clinical text. The ablation set spans MedGemma \cite{yang2024medgemma}, \ac{BERTbase} \cite{devlin2019bert}, \ac{GPTtwo} \cite{radford2019language}, and BioBART \cite{yuan2022biobart}. Under a shared protocol, we test whether report quality is more sensitive to frontend/interface design or to decoder family choice.

\section{Contributions}
The main contributions of this thesis are as follows:
\begin{enumerate}
    \item \textbf{fVLM-Based Report Generation Framework}: We build an end-to-end \ac{CT} report generation pipeline around a pretrained fVLM vision encoder, with a shared organ-aware visual frontend, decoder-side adaptation, and a reproducible \ac{BERTbase} reference branch.
    \item \textbf{Systematic Quality-Improvement Study}: We evaluate practical strategies to improve report generation quality, including organ-conditioned token interfaces, decoder choices such as MedGemma, \ac{BERTbase}, \ac{GPTtwo}, and BioBART, and training ablations under a controlled setup.
    \item \textbf{Clinical Efficacy Prioritization}: We validate generated results strictly against clinical-factuality metrics such as \ac{GREEN} and \ac{RadGraph} to distinguish genuine clinical correctness from superficial overlap-based gains driven by memorized reporting patterns.
\end{enumerate}

\section{Thesis Outline}
The remainder of this thesis is organized as follows. Chapter~\ref{chapter:related_work} reviews relevant literature and motivates the design choices used in this work. Chapter~\ref{chapter:methodology} describes the end-to-end pipeline and training setup. Chapter~\ref{chapter:experiments} presents the experimental results, Chapter~\ref{chapter:discussion} analyzes their implications and limitations, and Chapter~\ref{chapter:conclusion} summarizes the main outcomes and future directions.
