% !TeX root = ../main.tex
\chapter{Methodology}\label{chapter:methodology}

\section{Problem Formulation and Pipeline Overview}\label{sec:method_overview}
This thesis studies organ-aware report generation from 3D chest CT. The model input is a CT volume
$V \in \mathbb{R}^{D \times H \times W}$, together with organ masks
$\{M_o\}_{o \in \mathcal{O}}$ for a fixed organ set $\mathcal{O}$. The output is a structured report where each organ receives a dedicated text segment, and all segments are assembled into a final report for evaluation.

Formally, for patient $i$, we model:
\begin{equation}
\hat{Y}_i = f_\theta\!\left(V_i, \{M_{i,o}\}_{o \in \mathcal{O}}\right),
\end{equation}
where $\hat{Y}_i = \{\hat{y}_{i,o}\}_{o \in \mathcal{O}}$ contains organ-level generated text. We then build a report-level output by concatenating organ sections in a fixed anatomical order.

The end-to-end pipeline contains six stages:
\begin{enumerate}
    \item CT preprocessing and mask preparation.
    \item LLM-based decomposition of reference reports into organ-level supervision.
    \item Organ-aware visual token extraction from a 3D ViT backbone.
    \item Vision-language bridging into decoder embedding space.
    \item Organ-conditioned autoregressive generation.
    \item Organ-section assembly and report-level evaluation.
\end{enumerate}

The same overall interface is used for three decoder families: MedGemma~\cite{yang2024medgemma}, GPT-2~\cite{radford2019language}, and BioBART~\cite{yuan2022biobart}. This keeps decoder comparison meaningful by controlling visual input structure and data protocol.

\section{Data Interfaces and Preprocessing}\label{sec:method_preprocess}
\subsection{Volume and Mask Preprocessing}
The training and evaluation code uses a consistent MONAI-based transform stack across model families. Given raw NIfTI CT and segmentation volumes, preprocessing applies:
\begin{enumerate}
    \item channel standardization and axis transposition,
    \item HU intensity clipping ($[-1150, 350]$) and min-max scaling to $[0,1]$,
    \item spatial padding to $(112,256,352)$,
    \item final resize to $(112,256,352)$ with trilinear interpolation for CT and nearest-neighbor for masks.
\end{enumerate}

This choice intentionally replaces center-cropping with resizing to reduce accidental removal of peripheral anatomy (especially small or partially visible structures).

\subsection{Target Organs and Segmentation-ID Mapping}
The final training schema uses 12 target organs that are sufficiently represented in the decomposed reports and compatible with available segmentation labels:
\begin{equation}
\mathcal{O} = \{\text{lung, heart, esophagus, liver, gallbladder, stomach, pancreas, spleen, kidney, aorta, trachea, rib}\}.
\end{equation}

Mask extraction in data loaders follows fixed integer-ID mappings. Table~\ref{tab:organ_id_mapping} summarizes the mapping used in training and evaluation scripts.

\begin{table}[htbp]
    \centering
    \caption{Organ-to-segmentation-ID mapping used for binary mask construction.}
    \label{tab:organ_id_mapping}
    \begin{tabular}{ll}
        \toprule
        \textbf{Organ} & \textbf{Segmentation IDs} \\\midrule
        Lung & 10, 11, 12, 13, 14 \\
        Heart & 51, 61 \\
        Aorta & 52 \\
        Esophagus & 15 \\
        Trachea & 16 \\
        Rib & 92--115 \\
        Liver & 5 \\
        Gallbladder & 4 \\
        Stomach & 6 \\
        Pancreas & 7 \\
        Spleen & 1 \\
        Kidney & 2, 3 \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Chest-CT Visibility Caveat}
Although the task is chest CT, several upper-abdominal organs are retained because they appear in a non-trivial fraction of scans. However, visibility is inconsistent across subjects and protocols. Therefore, organs with limited field-of-view support are treated as optional supervision targets rather than mandatory textual outputs.

\section{Report Decomposition for Organ-Level Supervision}\label{sec:method_decomposition}
\subsection{LLM-Based Structuring}
Reference reports are decomposed into organ-keyed JSON records using an instruction-following LLM pipeline (Llama-3.3-70B-Instruct through vLLM~\cite{kwon2023efficient}, with AWQ quantization~\cite{lin2023awq}). The pipeline enforces:
\begin{enumerate}
    \item strict schema-constrained JSON output,
    \item one textual value per organ key,
    \item explicit handling of missing organs as empty/null values.
\end{enumerate}

This decomposition converts free-text report supervision into organ-conditioned generation targets and enables direct organ-level error analysis.

\subsection{Prompted Supervision Format}
During training, each organ is paired with an organ-specific instruction prompt. In MedGemma branches, chat-format prompts are used:
\begin{quote}
\ttfamily <start\_of\_turn>user ... Describe the findings for the [organ] ... <start\_of\_turn>model
\end{quote}
For GPT-2/BioBART branches, a shorter prompt style is used (e.g., \texttt{Describe [organ]:}).

The loss is masked to the response region only; prompt tokens and padding tokens are ignored.

\section{Organ-Aware Visual Encoding}\label{sec:method_visual_encoder}
\subsection{Backbone}
All main variants use a 3D ViT encoder initialized from a pretrained checkpoint and fine-tuned end-to-end. The standard configuration operates on $(112,256,352)$ volumes with patch size $(16,16,32)$ and hidden size 768.

Some ablations modify this backbone (e.g., CNN stem before ViT, multiscale fusion), but the output interface to the language side remains organ-conditioned tokens.

\subsection{Masked Cross-Attention with Organ Queries}
Let $F(V) \in \mathbb{R}^{T \times d_v}$ be ViT features and $Q_o \in \mathbb{R}^{K \times d_v}$ be $K$ learned queries for organ $o$. Binary organ masks are downsampled to the ViT token grid and converted to an additive attention bias matrix $B_o$, where background tokens receive $-\infty$.

Organ token extraction uses masked cross-attention:
\begin{equation}
Z_o = \mathrm{MHA}(Q_o, F(V), F(V); B_o),
\end{equation}
where $Z_o \in \mathbb{R}^{K \times d_v}$ and $K$ is the token count per organ.

In the implementation, the downsampled mask size is fixed to $(7,16,11)$ to match the ViT tokenization geometry used in training.

\subsection{Token Granularity: Multi-Token vs Compressed Tokens}
A major design axis is token granularity per organ:
\begin{itemize}
    \item \textbf{Multi-token mode:} $K=8$ visual tokens per organ (96 total for 12 organs).
    \item \textbf{Compressed mode:} one token per organ after aggregation/projection (effective $K=1$ at decoder interface).
\end{itemize}

The compressed strategy is motivated by 3D token-budget constraints and empirical instability when long visual prefixes are combined with long text prompts.

\subsection{Mask Safety and Empty-Organ Handling}
A practical issue in organ-conditioned attention is empty masks. If all attention positions are blocked (all $-\infty$), softmax becomes undefined. The implementation detects this case and resets the bias to zeros for that organ-query path, allowing fallback to unconstrained attention rather than producing NaNs.

\section{Architecture Variants Used in Ablations}\label{sec:method_arch_variants}
\subsection{Baseline: ROI Cross-Attention + Projector}
The baseline branch uses one masked cross-attention stage over ViT features and a linear projector into decoder embedding space.

\subsection{Positional Visual Prefix Variant}
The best-performing MedGemma variant adds learned visual positional embeddings and explicit sentinel wrapping (\texttt{<vis>} and \texttt{<end\_vis>}) around visual tokens before text tokens. This helps the decoder separate visual context from instruction text.

\subsection{Perceiver Resampler Variant}
A Perceiver-style resampler~\cite{jaegle2021perceiver} replaces single-step cross-attention with multiple layers, each containing:
\begin{enumerate}
    \item self-attention among organ queries,
    \item cross-attention from queries to ViT features,
    \item feed-forward refinement.
\end{enumerate}
This increases modeling capacity but also compute and optimization complexity.

\subsection{Multi-Scale ViT-FPN Variant}
The multi-scale variant fuses intermediate ViT block features (early/mid/late/final) using learned scale weights and a fusion projection. The goal is to preserve both coarse and fine signals, especially for small or low-contrast structures.

\subsection{MedGemma Architecture v3 Variant}
The v3 branch introduces a CNN stem before ViT, changes the ViT patch geometry accordingly, and adds a text-visual alignment term during training. This variant explores whether stronger inductive bias in the encoder and auxiliary contrastive alignment improves organ grounding.

\section{Vision-Language Bridging}\label{sec:method_bridge}
\subsection{Projection and Stabilization}
Visual features are projected from ViT dimension $d_v=768$ to decoder hidden size $d_l$:
\begin{equation}
\tilde{Z}_o = W_p Z_o + b_p, \quad W_p \in \mathbb{R}^{d_v \times d_l}.
\end{equation}

Two stabilization operations are applied before decoding:
\begin{enumerate}
    \item layer normalization on projected visual embeddings,
    \item variance matching to decoder token embedding scale.
\end{enumerate}

The variance matching step reduces ``embedding shock'' when concatenating projected visual vectors with pretrained language embeddings.

\subsection{Sentinel Wrapping and Token Assembly}
For MedGemma, the effective input sequence is:
\begin{equation}
[\text{BOS}],\;\texttt{<vis>},\;\tilde{Z}_o,\;\texttt{<end\_vis>},\;\text{text prompt and target tokens}.
\end{equation}

Loss is masked over the inserted visual block so that only clinically relevant text tokens contribute to language-modeling gradients.

\section{Decoder-Specific Adaptation}\label{sec:method_decoders}
\subsection{MedGemma Branch}
MedGemma is loaded as a frozen base decoder, then adapted with LoRA~\cite{hu2022lora}. LoRA settings used in main runs are:
\begin{itemize}
    \item rank $r=16$,
    \item $\alpha=32$,
    \item dropout $0.05$,
    \item target modules: \texttt{q\_proj, k\_proj, v\_proj, o\_proj, gate\_proj, up\_proj, down\_proj}.
\end{itemize}

This keeps training memory manageable while allowing meaningful decoder adaptation under organ-conditioned visual inputs.

\subsection{GPT-2 and BioBART Branches}
For GPT-2/BioBART, the model is wrapped in a \texttt{VisionEncoderDecoderModel} interface. The encoder output is organ-token embeddings; decoder-side adaptation unfreezes a restricted subset of parameters (cross-attention blocks, normalization layers, output head, and embeddings) while keeping most pretrained weights frozen.

For seq2seq decoders (BioBART), pretrained shared embeddings and output head weights are explicitly copied from the original seq2seq checkpoint to avoid random initialization artifacts in causal-decoding wrappers.

\section{Training Objectives}\label{sec:method_objectives}
\subsection{Weighted Language Modeling Loss}
Given batch index $i$, organ $o$, and token position $t$, the primary loss is weighted token-level cross-entropy:
\begin{equation}
\mathcal{L}_{\mathrm{LM}} = \frac{1}{B|\mathcal{O}|} \sum_{i=1}^{B}\sum_{o\in\mathcal{O}} w_{i,o}
\left(\frac{1}{T_{i,o}}\sum_{t=1}^{T_{i,o}} \mathrm{CE}(y_{i,o,t}, \hat{y}_{i,o,t})\right),
\end{equation}
where $w_{i,o}$ is an organ-sample weight used for imbalance handling.

The weighting mechanism differs across ablations:
\begin{itemize}
    \item probability-based undersampling of default ``no finding'' targets,
    \item explicit organ loss weights from precomputed statistics,
    \item curriculum or hard-example dynamic weights.
\end{itemize}

\subsection{Optional Alignment Loss}
Some variants (notably \texttt{medgemma\_architecture\_v3} and GPT-2/BioBART branches) add a contrastive alignment objective between organ visual and organ text representations. For patient-level organ sets, similarity logits are computed as:
\begin{equation}
S_{o,o'} = \frac{\langle v_o, t_{o'} \rangle}{\tau},
\end{equation}
followed by cross-entropy over matching organ indices.

The total objective becomes:
\begin{equation}
\mathcal{L} = \mathcal{L}_{\mathrm{LM}} + \lambda \mathcal{L}_{\mathrm{align}},
\end{equation}
with $\lambda$ set by variant-specific training code.

\section{Optimization and Training Protocol}\label{sec:method_training}
\subsection{Shared Optimization Settings}
Across major runs, training arguments are largely standardized:
\begin{itemize}
    \item batch size: 1 per device,
    \item gradient accumulation: 8,
    \item learning rate: $1\times10^{-4}$,
    \item weight decay: 0.01,
    \item warmup ratio: 0.05,
    \item epochs: 2 (retraining phase),
    \item bf16 enabled, fp16 disabled,
    \item gradient checkpointing enabled,
    \item checkpoint/evaluation interval: 200 steps.
\end{itemize}

The low per-device batch size is a practical memory constraint of 3D volumes and large decoders; gradient accumulation is therefore critical for stable optimization.

\subsection{Training-Strategy Ablations}
Beyond architecture, several training-strategy variants are implemented:
\begin{itemize}
    \item \textbf{Curriculum learning:} first phase focuses on pathological organ samples, then switches to full distribution.
    \item \textbf{Hard example mining:} dynamically upsamples high-loss samples using weighted random sampling.
    \item \textbf{Undersampling default text:} reduces dominance of ``normal'' templates.
    \item \textbf{Organ-weighted loss:} separate weights for explicit findings and default findings per organ.
\end{itemize}

These ablations test whether optimization strategy can compensate for class imbalance in organ-conditioned report targets.

\section{Inference and Report Assembly}\label{sec:method_inference}
\subsection{Organ-wise Generation}
At inference, one prompt is created per organ, and generation runs in organ-parallel mode within each patient. Decoding settings are model-family specific:
\begin{itemize}
    \item \textbf{MedGemma runs:} beam search (typically 3 beams), \texttt{max\_new\_tokens=100}, \texttt{no\_repeat\_ngram\_size=3}, moderate repetition penalty.
    \item \textbf{GPT-2/BioBART runs:} beam search (typically 4 beams), \texttt{max\_length=120}, \texttt{no\_repeat\_ngram\_size=3}, stronger repetition penalty.
\end{itemize}

\subsection{Structured Report Reconstruction}
Generated organ segments are concatenated in a fixed order:
\begin{equation}
\hat{R} = \texttt{LUNG: }\hat{y}_{\mathrm{lung}}\;\Vert\;\cdots\;\Vert\;\texttt{KIDNEY: }\hat{y}_{\mathrm{kidney}}.
\end{equation}

Organs with empty/insufficient visual support are not forced to contain abnormal statements. This is important to avoid schema-driven hallucination where a model invents findings solely to populate every key.

\section{Model Comparison Matrix}\label{sec:method_model_matrix}
Table~\ref{tab:model_matrix_method} summarizes the major variants evaluated in this thesis.

\begin{table}[htbp]
    \centering
    \caption{Model variants and principal design differences.}
    \label{tab:model_matrix_method}
    \begin{tabular}{p{3.2cm}p{10.8cm}}
        \toprule
        \textbf{Variant} & \textbf{Main Design Choice} \\\midrule
        \texttt{medical\_vlm\_8\_tokens\_full} & Baseline MedGemma branch with 8 organ tokens, masked cross-attention, projector, and weighted LM training. \\
        \texttt{medgemma\_lora\_vis\_token\_pos\_embed} & Adds visual positional embeddings and explicit visual sentinel wrapping; strongest global performance in current runs. \\
        \texttt{perceiver\_resampler} & Replaces single cross-attention with stacked Perceiver-style self/cross-attention blocks. \\
        \texttt{multiscale\_vit\_fpn} & Fuses intermediate ViT features from multiple depths before organ-query attention. \\
        \texttt{medgemma\_architecture\_v3} & Adds CNN stem and visual-text alignment loss; explores stronger encoder bias and contrastive regularization. \\
        \texttt{curriculum\_learning} & Two-phase pathology-first curriculum before full distribution training. \\
        \texttt{hard\_example\_mining} & Dynamic weighted sampler emphasizing high-loss samples. \\
        GPT-2 / BioBART branches & Same organ-attention encoder interface with decoder-family adaptation for cross-family comparison. \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Reproducibility and Implementation Notes}\label{sec:method_reproducibility}
\begin{itemize}
    \item Data, training, and evaluation are organized per-variant in dedicated folders to avoid cross-contamination of checkpoints and metrics.
    \item Each official run stores generated reports and a standardized \texttt{metrics\_final.csv} summary for reproducible chapter tables.
    \item Metrics are computed through a unified RadEval-based script with fixed key mapping into thesis metric names.
    \item Decoder comparisons are restricted to official organ-attention result folders to prevent mixing with earlier exploratory runs.
\end{itemize}

This methodological design keeps the contribution focused: controlling 3D token budget and organ grounding while comparing architecture and decoder choices under one consistent data protocol.
