\chapter{Experiments and Results}\label{chapter:experiments}

\section{Dataset}
We validate all models on \textbf{CT-RATE}~\cite{hamamci2024foundation}, a large multimodal dataset for 3D CT. CT-RATE pairs volumetric chest CT scans with free-text reports and provides an established benchmark setting for report generation and related multimodal tasks~\cite{hamamci2024generatect}.

\subsection{Dataset Overview}
The dataset contains 25,692 CT volumes from 21,304 unique patients. We follow the official train/validation/test split.

\begin{table}[htbp]
    \centering
    \caption{CT-RATE statistics used in this thesis.}
    \label{tab:ctrate_stats}
    \begin{tabular}{ll}
        \toprule
        \textbf{Statistic} & \textbf{Value} \\ \midrule
        Total CT Volumes & 25,692 \\
        Unique Patients & 21,304 \\
        Modality & 3D Chest CT \\
        Text Modality & Findings + Impression \\ 
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Preprocessing}
We use a fixed preprocessing pipeline:
\begin{enumerate}
    \item Intensity clipping and normalization in HU space.
    \item Spatial normalization to a fixed volume shape.
    \item Channel-first tensor layout for 3D encoder input.
    \item LLM-based decomposition of reports into organ-specific text segments.
\end{enumerate}

We use decomposed organ-level supervision for training and standard report-level reconstruction for evaluation.

\section{Evaluation Metrics}\label{sec:evaluation_metrics}
To assess the performance of our radiology report generation system we evaluate generated reports with multiple complementary metrics. Metrics are grouped by the type of signal they capture:

\begin{enumerate}
    \item \textbf{Surface-level (linguistic) metrics:} measure n-gram overlap and surface similarity to reference reports.
    \item \textbf{Semantic / embedding-based metrics:} measure semantic similarity using contextual embeddings.
    \item \textbf{Clinical / structured metrics:} evaluate clinical correctness by extracting findings, entities, relations or clinically-relevant labels.
    \item \textbf{Human evaluation and statistical tests:} measure radiologist preference, clinical usefulness, and statistical significance.
\end{enumerate}

\subsection{Surface-level (Natural Language Generation) Metrics}
These metrics are commonly used in image captioning and report generation benchmarks and capture lexical and syntactic similarity.

\subsubsection{BLEU}
BLEU \cite{papineni2002bleu} is an n-gram precision-based metric. For a candidate report $c$ and a set of references $\{r\}$, BLEU is computed as
\begin{equation}
    \text{BLEU} = \text{BP} \cdot \exp\!\left(\sum_{n=1}^{N} w_n \log p_n \right),
\end{equation}
where $p_n$ is the modified precision for $n$-grams, $w_n$ are weights (we use uniform weights $w_n = 1/N$), and BP is the brevity penalty:
\begin{equation}
    \text{BP} =
    \begin{cases}
        1 & \text{if } c_{len} > r_{len}^* \\
        \exp\left(1 - \frac{r_{len}^*}{c_{len}} \right) & \text{otherwise}
    \end{cases}
\end{equation}
Here $c_{len}$ is the candidate length and $r_{len}^*$ is the effective reference length (e.g., closest reference length). We report BLEU-4.

\subsubsection{ROUGE-L}
ROUGE-L \cite{lin2004rouge} uses the length of the Longest Common Subsequence (LCS) to capture sentence-level structure. Define
\[
R_{LCS} = \frac{\text{LCS}(c, r)}{|r|}, \qquad P_{LCS} = \frac{\text{LCS}(c, r)}{|c|},
\]
then the ROUGE-L F-measure is
\begin{equation}
F_{LCS} = \frac{(1+\beta^2) \, R_{LCS} \, P_{LCS}}{R_{LCS} + \beta^2 P_{LCS}},
\end{equation}
where $\beta$ weights recall relative to precision. In practice we set $\beta=1$ to report the balanced F$_1$ unless otherwise specified.

\subsubsection{METEOR}
METEOR \cite{banerjee2005meteor} aligns unigrams and computes a weighted harmonic mean of precision and recall that emphasizes recall. The core F$_{mean}$ is:
\begin{equation}
    F_{\text{mean}} = \frac{(1+\alpha) P R}{R + \alpha P},
\end{equation}
where typical METEOR uses $\alpha=9$ (equivalently $F_{\text{mean}} = \frac{10PR}{R + 9P}$). METEOR applies a fragmentation penalty:
\begin{equation}
    \text{Penalty} = 0.5 \left(\frac{\#\text{chunks}}{\#\text{matches}}\right)^3,
\end{equation}
and final score:
\begin{equation}
    \text{METEOR} = F_{\text{mean}} \cdot (1 - \text{Penalty}).
\end{equation}

\subsubsection{CIDEr}
CIDEr \cite{vedantam2015cider} uses TF–IDF weighting for n-grams and computes average cosine similarity between a candidate's and references' TF–IDF vectors. For n-grams of length $n$:
\begin{equation}
    \text{CIDEr}_n(c, r) = \frac{1}{M}\sum_{i=1}^{M}
    \frac{\mathbf{g}^n(c)\cdot \mathbf{g}^n(r_i)}
             {\|\mathbf{g}^n(c)\| \, \|\mathbf{g}^n(r_i)\|},
\end{equation}
and final CIDEr is the (optionally IDF-normalized) average across $n=1\ldots4$.

\vspace{2mm}
\textbf{Notes:} Surface metrics are sensitive to wording and may penalize clinically-correct paraphrases. Use them together with semantic and clinical metrics.

\subsection{Semantic / Embedding-based Metrics}
These metrics capture deeper (contextual) semantic similarity beyond n-gram overlap.

\subsubsection{BERTScore}
BERTScore~\cite{zhang2019bertscore} compares contextual token embeddings between generated and reference reports and aggregates token-level similarities into precision, recall, and F1-style semantic similarity.

\subsection{Clinical / Structured Metrics}
These metrics aim to measure clinical correctness and the presence/absence of findings.

\subsubsection{CheXbert}
CheXbert \cite{smit2020chexbert} is a BERT-based labeler trained on the output of the CheXpert rule-based labeler to extract findings from radiology reports with higher accuracy. We use it to detect the presence of 14 standard observations (e.g., Cardiomegaly, Edema, Pneumonia, No Finding). We compute the Micro-F$_1$ score across all 14 classes to measure the clinical accuracy of the generated reports compared to the ground truth.

\subsubsection{RadGraph}
RadGraph \cite{jain2021radgraph} converts radiology reports into knowledge graphs (entities and relations). Evaluation is done by measuring overlap between predicted and reference graphs; typical metrics are node (entity) Precision/Recall/F$_1$ and edge (relation) Precision/Recall/F$_1$. Reporting both entity-level and relation-level scores helps diagnose whether models miss or hallucinate links between findings and anatomy.

\subsubsection{SRR-BERT / Structured-report metrics}
SRR-BERT and similar methods fine-tune BERT-like encoders to score structured-report correctness; they are useful for fine-grained disease classification and structured-report components. Report the metric(s) as provided by the original implementation (precision/recall/F$_1$ or regression correlation).

\subsubsection{GREEN and other factuality metrics}
GREEN (Generative Radiology report Evaluation and Error Notation) and related metrics are designed to highlight factual errors and categorize them by severity (critical vs non-critical). Use them to count clinically-significant hallucinations and factual inconsistencies.

\subsubsection{Composite clinical metrics (e.g., RadCliQ)}
Composite metrics such as RadCliQ combine surface (e.g., BLEU) and clinical (e.g., CheXpert) signals, often with learned weights to better correlate with radiologist judgments. When using a learned composite, report how weights were obtained and validate correlation with human raters.

\section{Results}
\subsection{A. Internal MedGemma Ablations (CT-RATE)}
Table~\ref{tab:medgemma_ablation} reports core MedGemma-family ablations using the official aggregated results from \texttt{fvlm/rep\_medgemma/results\_retrain/global\_model\_comparison.csv}.

\begin{table}[htbp]
    \centering
    \caption{Internal MedGemma ablations on CT-RATE test set (GLOBAL metrics).}
    \label{tab:medgemma_ablation}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{lcccccccccc}
        \toprule
        \textbf{Model} & \textbf{GREEN} & \textbf{BLEU-4} & \textbf{METEOR} & \textbf{ROUGE-L} & \textbf{CIDEr} & \textbf{BERTScore} & \textbf{RadGraph} & \textbf{CheXbert} & \textbf{RadCliQ} & \textbf{SRR-BERT} \\ \midrule
        medical\_vlm\_8\_tokens\_full (baseline) & 0.323 & 0.092 & 0.328 & 0.341 & 0.041 & 0.567 & 0.243 & 0.239 & 1.037 & 0.330 \\
        medgemma\_lora\_vis\_token\_pos\_embed (best) & \textbf{0.348} & \textbf{0.105} & \textbf{0.344} & \textbf{0.379} & \textbf{0.084} & \textbf{0.596} & \textbf{0.268} & 0.026 & \textbf{1.104} & 0.324 \\
        multiscale\_vit\_fpn & 0.327 & 0.086 & 0.322 & 0.337 & 0.043 & 0.566 & 0.242 & 0.241 & 1.022 & 0.326 \\
        perceiver\_resampler & 0.316 & 0.082 & 0.317 & 0.333 & 0.036 & 0.564 & 0.236 & 0.221 & 1.008 & 0.308 \\
        medgemma\_architecture\_v3 & 0.313 & 0.083 & 0.318 & 0.335 & 0.030 & 0.564 & 0.242 & 0.206 & 1.007 & 0.301 \\
        curriculum\_learning & 0.194 & 0.032 & 0.250 & 0.250 & 0.018 & 0.507 & 0.133 & 0.269 & 0.773 & 0.322 \\
        \bottomrule
    \end{tabular}
    }
\end{table}

The strongest result is obtained by the single-token organ representation with positional embedding, while curriculum training performs worst among tested variants.

\subsection{B. Decoder Comparison (CT-RATE, Core Axis)}
Decoder-family comparison is a central part of this thesis. Table~\ref{tab:decoder_comparison} is the final comparison template used for MedGemma, GPT-2, and BioBART under the same organ-attention pipeline.

\begin{table}[htbp]
    \centering
    \caption{Decoder comparison on CT-RATE. GPT-2/BioBART rows will be filled from the official \texttt{rep\_vision\_organ\_attention} runs once \texttt{metrics\_final.csv} is generated.}
    \label{tab:decoder_comparison}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{lcccccccccc}
        \toprule
        \textbf{Decoder} & \textbf{GREEN} & \textbf{BLEU-4} & \textbf{METEOR} & \textbf{ROUGE-L} & \textbf{CIDEr} & \textbf{BERTScore} & \textbf{RadGraph} & \textbf{CheXbert} & \textbf{RadCliQ} & \textbf{SRR-BERT} \\ \midrule
        MedGemma (best variant) & 0.348 & 0.105 & 0.344 & 0.379 & 0.084 & 0.596 & 0.268 & 0.026 & 1.104 & 0.324 \\
        GPT-2 (rep\_vision\_organ\_attention/results/gpt2\_v3) & -- & -- & -- & -- & -- & -- & -- & -- & -- & -- \\
        BioBART (rep\_vision\_organ\_attention/results/biobart\_v3) & -- & -- & -- & -- & -- & -- & -- & -- & -- & -- \\
        \bottomrule
    \end{tabular}
    }
\end{table}

To keep the comparison protocol consistent, GPT-2 and BioBART metrics are taken only from the official organ-attention evaluation outputs (\texttt{generated\_reports.csv} + \texttt{metrics\_final.csv}) and not from older fragmented experiment folders.

\subsection{C. External fVLM Baseline (Contextual, Not Directly Comparable)}
We additionally report fVLM~\cite{shui2025fvlm} as an external baseline reference. This baseline is contextual only: training data, preprocessing, and evaluation protocol differ from our internal CT-RATE setup.

\begin{table}[htbp]
    \centering
    \caption{External baseline reference. Values should be read as paper-reported context, not direct row-wise comparison with internal CT-RATE runs.}
    \label{tab:external_fvlm}
    \begin{tabular}{p{2.6cm}p{3.2cm}p{6.2cm}}
        \toprule
        \textbf{Model} & \textbf{Protocol} & \textbf{Interpretation in this thesis} \\ \midrule
        fVLM~\cite{shui2025fvlm} & Large-scale fine-grained CT VLM pretraining (paper setting) & Used as external reference point for report-generation performance trends and organ-aware supervision design. \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Discussion of Metric Behavior}
During internal ablations, we observe that CheXbert can be unstable for organ-decomposed CT reports, consistent with the fact that CheXbert was designed for full chest-report style labeling. Therefore, we prioritize GREEN and RadGraph when interpreting clinical faithfulness in this organ-conditioned setting.
